---
title: "Cost-Effectiveness Analysis"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
gradethis_setup()
knitr::opts_chunk$set(echo = FALSE)

tutorial_options(exercise.eval = TRUE)
```

## Introduction

Over the last few weeks, we have defined some model input parameters and code to implement a cohort-level Markov model and an individual-level state transition model. The code for these is below (which should be similar to your previous scripts -- I've revised a few of the values slightly).

(These are interactive code boxes - you can play around with these to change the code as you like. Hit 'Run' to run your code and see the output (if it produces any), and 'Start Over' to reset the code back to it's original state.)

```{r parameters, exercise=TRUE}
states <- c("healthy", "acl_acute", "acl_chronic", "acl_recovered", "oa", "tkr", "post_tkr", "dead")

probs <- c(0, 1, 0, 0, 0, 0, 0, 0)
names(probs) <- states

transitions <- rbind(c(1, 0, 0,    0,   0,    0,   0,    0),
                     c(0, 0, 0.44, 0.5, 0.05, 0,   0,    0.01),
                     c(0, 0, 0.38, 0.5, 0.1,  0,   0,    0.02),
                     c(0, 0, 0,    0.9, 0.08, 0,   0,    0.02),
                     c(0, 0, 0,    0,   0.87, 0.1, 0,    0.03),
                     c(0, 0, 0,    0,   0,    0,   0.96, 0.04),
                     c(0, 0, 0,    0,   0,    0,   0.95, 0.05),
                     c(0, 0, 0,    0,   0,    0,   0,    1))
rownames(transitions) <- colnames(transitions) <- states

costs <- c(0, 7000, 1200, 500, 1000, 20000, 1000, 0)
names(costs) <- states
utility <- c(1, 0.5, 0.8, 0.9, 0.7, 0.6, 0.8, 0)
names(utility) <- states
```

```{r cohort-model, exercise=TRUE, exercise.setup = "parameters"}
update_probabilities <- function(probs, transitions) {
  as.vector(probs %*% transitions)
}

run_cohort_model <- function(initial_probs, transitions, t) {
  probabilities <- matrix(nrow = t + 1, ncol = length(initial_probs))
  probabilities[1, ] <- initial_probs

  for (x in seq_len(t)) {
    probabilities[x + 1, ] <- update_probabilities(probabilities[x, ],
                                                   transitions)
  }

  probabilities
}

calculate_outcome_value <- function(probs, values) {
  sum(probs * values)
}

calculate_outcomes <- function(probabilities, values, discount_rate = 0) {
  t <- nrow(probabilities)
  sum((probabilities %*% values) * (1 - discount_rate) ^ ((1:t) - 1))
}

```

```{r individual-model, exercise=TRUE, exercise.setup = "cohort-model"}
update_state <- function(initial_state, transitions) {
  n <- nrow(transitions)
  sample.int(n, 1, prob = transitions[initial_state, ])
}

simulate_individual <- function(initial_state, transitions, t) {
  states <- vector("integer", length = t + 1)
  states[1] <- initial_state

  for (x in seq_len(t)) {
    states[x + 1] <- update_state(states[x], transitions)
  }

  states
}

run_individual_model <- function(initial_states, transitions, t) {
  n <- length(initial_states)

  states <- matrix(nrow = n, ncol = t + 1)
  for (x in seq_len(n)) {
    states[x, ] <- simulate_individual(initial_states[x], transitions, t)
  }

  states
}

calculate_individual_outcome_value <- function(states, values) {
  matrix(values[states], nrow = nrow(states))
}

calculate_average_outcomes <- function(states, values, discount_rate = 0) {
  outcomes <- calculate_individual_outcome_value(states, values)
  n <- nrow(outcomes)
  t <- ncol(outcomes)
  for (x in seq_len(n)) {
    outcomes[x, ] <- outcomes[x, ] * (1 - discount_rate) ^ ((1:t) - 1)
  }
  mean(rowSums(outcomes))
}

```

## Cost-effectiveness analysis

To use these models for cost-effectiveness analysis, we need to define model input parameters that represent each of the interventions to be compared -- these could involve different values for transition probabilities, costs, and/or outcomes (health utility).

For a real analysis, these would be estimated from data, such as a randomised trial, observational study, or systematic review/meta-analysis; for this example we will just make up some numbers :)

### Usual care comparator
Let's assume our previous input parameters represent 'business-as-usual', or an appropriate usual care comparator for the economic evaluation:

```{r transitions_uc, exercise=TRUE, exercise.setup = "individual-model"}
transitions_UC <- transitions
transitions_UC
```

```{r costs_uc, exercise=TRUE, exercise.setup = "transitions_uc"}
costs_UC <- costs
costs_UC
```

```{r utility_uc, exercise=TRUE, exercise.setup = "costs_uc"}
utility_UC <- utility
utility_UC
```

```{r params_uc, exercise=TRUE, exercise.setup = "utility_uc"}
params_UC <- list(transitions = transitions_UC, costs = costs_UC, utility = utility_UC)
```

### New interventions to evaluate:

We'll consider a few hypothetical interventions to try in our model.

First, an intervention that improves symptomatic recovery from ACL injury (reduces the probability of transition from `acl_acute` to `acl_chronic` (symptomatic) and increases the probability of transition to `acl_recovered` (asymptomatic)), but costs an additional $2000 per patient (over the year following injury):

```{r improved-recovery-parameters, exercise=TRUE, exercise.setup = "params_uc"}
transitions_IR <- transitions
transitions_IR[2, ] <- c(0, 0, 0.24, 0.7, 0.05, 0, 0, 0.01)

costs_IR <- costs
costs_IR[2] <- 9000

params_IR <- list(transitions = transitions_IR, costs = costs_IR, utility = utility_UC)
```

Another that provides ongoing treatment to reduce the utility losses in the post-ACL states:

```{r ongoing-treatment-parameters, exercise=TRUE, exercise.setup = "improved-recovery-parameters"}
utility_OT <- utility
utility_OT[3:4] <- c(0.85, 0.95)

costs_OT <- costs
costs_OT[3:4] <- c(1700, 1000)

params_OT <- list(transitions = transitions_UC, costs = costs_OT, utility = utility_OT)
```

And another that reduces the rate of OA incidence from the post-ACL states:

```{r reduced-oa-incidence-parameters, exercise=TRUE, exercise.setup = "ongoing-treatment-parameters"}
transitions_RO <- transitions
transitions_RO[3, 3:5] <- c(0.43, 0.5, 0.05)
transitions_RO[4, 4:5] <- c(0.94, 0.04)

costs_RO <- costs
costs_RO[3:4] <- c(1500, 1000)

params_RO <- list(transitions = transitions_RO, costs = costs_RO, utility = utility_UC)
```

## CEA (cont.)

###

To complete our analysis, we will want to write some functions to (1) run the simulation model for each set of input parameters; (2) calculate the resulting cohort outcomes (i.e. aggregate costs and QALYs); and (3) compare the outcomes between interventions to compute cost-effectiveness. Let's start with the cohort-level model.

### Simulation model function

First, we want a function that runs the cohort simulation model twice: once for the comparator scenario (`UC`, for usual care) and once for the intervention of interest (`TX`); and returns the two sets of cohort outcomes.

These next code boxes require you to fill in the gaps with the correct code fragments. Use 'Run code' to run the code and see the result, and 'Submit answer' to check if you've completed it correctly. If you get stuck, the 'Solution' button will show you my solution. If you mess everything up, try 'Start Over'!

```{r cohort-simulation-function, exercise=TRUE, exercise.eval=FALSE, exercise.setup="reduced-oa-incidence-parameters", exercise.lines=32}
cohort_simulation <- function(initial_probs, parameters_UC, parameters_TX, horizon, discount_rate) {
  probabilities_UC <- run_cohort_model(initial_probs = ____,
                                       transitions = ____,
                                       t = ____)
  probabilities_TX <- run_cohort_model(initial_probs = ____,
                                       transitions = ____,
                                       t = ____)
  
  Costs_UC <- calculate_outcomes(probabilities = ____,
                                 values = ____,
                                 discount_rate = ____)
  Costs_TX <- calculate_outcomes(probabilities = ____,
                                 values = ____,
                                 discount_rate = ____)
  
  QALYs_UC <- calculate_outcomes(probabilities = ____,
                                 values = ____,
                                 discount_rate = ____)
  QALYs_TX <- calculate_outcomes(probabilities = ____,
                                 values = ____,
                                 discount_rate = ____)
  
  list(UC = list(Costs = Costs_UC, QALYs = QALYs_UC),
       TX = list(Costs = Costs_TX, QALYs = QALYs_TX))
}

# Let's test that out with our 'Improved Recovery' intervention:
simulation_IR <- cohort_simulation(
  probs, parameters_UC = params_UC, parameters_TX = params_IR,
  horizon = 10, discount_rate = 0.035
)
simulation_IR
```

```{r cohort-simulation-function-hint-1}
probabilities_UC <- run_cohort_model(initial_probs = initial_probs,
                                     transitions = parameters_UC$transitions,
                                     t = horizon)
```

```{r cohort-simulation-function-hint-2}
probabilities_UC <- run_cohort_model(initial_probs = initial_probs,
                                     transitions = parameters_UC$transitions,
                                     t = horizon)

...

Costs_UC <- calculate_outcomes(probabilities = probabilities_UC,
                               values = parameters_UC$costs,
                               discount_rate = discount_rate)

...
```

```{r cohort-simulation-function-solution, exercise.reveal_solution=FALSE}
cohort_simulation <- function(initial_probs, parameters_UC, parameters_TX, horizon, discount_rate) {
  probabilities_UC <- run_cohort_model(initial_probs = initial_probs,
                                       transitions = parameters_UC$transitions,
                                       t = horizon)
  probabilities_TX <- run_cohort_model(initial_probs = initial_probs,
                                       transitions = parameters_TX$transitions,
                                       t = horizon)
  
  Costs_UC <- calculate_outcomes(probabilities = probabilities_UC,
                                 values = parameters_UC$costs,
                                 discount_rate = discount_rate)
  Costs_TX <- calculate_outcomes(probabilities = probabilities_TX,
                                 values = parameters_TX$costs,
                                 discount_rate = discount_rate)
  
  QALYs_UC <- calculate_outcomes(probabilities = probabilities_UC,
                                 values = parameters_UC$utility,
                                 discount_rate = discount_rate)
  QALYs_TX <- calculate_outcomes(probabilities = probabilities_TX,
                                 values = parameters_TX$utility,
                                 discount_rate = discount_rate)
  
  list(UC = list(Costs = Costs_UC, QALYs = QALYs_UC),
       TX = list(Costs = Costs_TX, QALYs = QALYs_TX))
}

# Let's test that out with our 'Improved Recovery' intervention:
simulation_IR <- cohort_simulation(
  probs, parameters_UC = params_UC, parameters_TX = params_IR,
  horizon = 10, discount_rate = 0.035
)
simulation_IR
```

```{r cohort-simulation-function-check}
grade_this_code()
```

```{r cohort-simulation-definition, exercise.setup="reduced-oa-incidence-parameters"}
cohort_simulation <- function(initial_probs, parameters_UC, parameters_TX, horizon, discount_rate) {
  probabilities_UC <- run_cohort_model(initial_probs, transitions = parameters_UC$transitions, t = horizon)
  probabilities_TX <- run_cohort_model(initial_probs, transitions = parameters_TX$transitions, t = horizon)
  
  Costs_UC <- calculate_outcomes(probabilities_UC, parameters_UC$costs, discount_rate = discount_rate)
  Costs_TX <- calculate_outcomes(probabilities_TX, parameters_TX$costs, discount_rate = discount_rate)
  
  QALYs_UC <- calculate_outcomes(probabilities_UC, parameters_UC$utility, discount_rate = discount_rate)
  QALYs_TX <- calculate_outcomes(probabilities_TX, parameters_TX$utility, discount_rate = discount_rate)
  
  list(UC = list(Costs = Costs_UC, QALYs = QALYs_UC),
       TX = list(Costs = Costs_TX, QALYs = QALYs_TX))
}

simulation_IR <- cohort_simulation(
  probs, parameters_UC = params_UC, parameters_TX = params_IR,
  horizon = 10, discount_rate = 0.035
)

```

### Cost-effectiveness outcomes

The next step is to calculate the incremental cost-effectiveness outcomes from these: let's just do the ICER, for now.

```{r icer-function, exercise=TRUE, exercise.eval=FALSE, exercise.setup="cohort-simulation-definition"}
calculate_ICER <- function(simulation_output) {
  incremental_costs <- simulation_output$____ - simulation_output$____
  incremental_qalys <- simulation_output$____ - simulation_output$____
  
  list(Costs = incremental_costs,
       QALYs = incremental_qalys,
       ICER = ____ / ____)
}

ICER_IR <- calculate_ICER(simulation_IR)
ICER_IR
```

```{r icer-function-hint-1}
incremental_costs <- simulation_output$TX$Costs - simulation_output$UC$Costs
```

```{r icer-function-solution, exercise.reveal_solution=FALSE}
calculate_ICER <- function(simulation_output) {
  incremental_costs <- simulation_output$TX$Costs - simulation_output$UC$Costs
  incremental_qalys <- simulation_output$TX$QALYs - simulation_output$UC$QALYs
  
  list(Costs = incremental_costs,
       QALYs = incremental_qalys,
       ICER = incremental_costs / incremental_qalys)
}

ICER_IR <- calculate_ICER(simulation_IR)
ICER_IR
```

```{r icer-function-check}
grade_this_code()
```

```{r icer-definition, exercise.setup="cohort-simulation-definition"}
calculate_ICER <- function(simulation_output) {
  incremental_costs <- simulation_output$TX$Costs - simulation_output$UC$Costs
  incremental_qalys <- simulation_output$TX$QALYs - simulation_output$UC$QALYs
  
  list(Costs = incremental_costs,
       QALYs = incremental_qalys,
       ICER = incremental_costs / incremental_qalys)
}

ICER_IR <- calculate_ICER(simulation_IR)
```

## Using our new CEA model

Now that we have a complete model, we can easily apply it to analyse other intervention scenarios.

The same model will work for the 'Ongoing Treatment' (reduced utility losses) intervention:

```{r cea-ot, exercise=TRUE, exercise.eval=FALSE, exercise.setup="icer-definition"}
simulation_OT <- cohort_simulation(
  initial_probs = ____,
  parameters_UC = ____,
  parameters_TX = ____,
  horizon = 10, discount_rate = 0.035
)
ICER_OT <- calculate_ICER(____)
ICER_OT
```

```{r cea-ot-hint-1}
# Recall how we did the simulation for the 'Improved Recovery' intervention:
simulation_IR <- cohort_simulation(
  initial_probs = probs,
  parameters_UC = params_UC,
  parameters_TX = params_IR,
  horizon = 10, discount_rate = 0.035
)
```

```{r cea-ot-solution, exercise.reveal_solution=FALSE}
simulation_OT <- cohort_simulation(
  initial_probs = probs,
  parameters_UC = params_UC,
  parameters_TX = params_OT,
  horizon = 10, discount_rate = 0.035
)
ICER_OT <- calculate_ICER(simulation_OT)
ICER_OT
```

```{r cea-ot-check}
grade_this_code()
```

```{r ot-cea-definition, exercise.setup="icer-definition"}
simulation_OT <- cohort_simulation(
  initial_probs = probs,
  parameters_UC = params_UC,
  parameters_TX = params_OT,
  horizon = 10, discount_rate = 0.035
)
ICER_OT <- calculate_ICER(simulation_OT)
```

And the 'Reduced OA incidence' intervention:

```{r cea-ro, exercise=TRUE, exercise.eval=FALSE, exercise.setup="ot-cea-definition"}
simulation_RO <- cohort_simulation(
  initial_probs = ____,
  parameters_UC = ____,
  parameters_TX = ____,
  horizon = 10, discount_rate = 0.035
)
ICER_RO <- calculate_ICER(____)
ICER_RO
```

```{r cea-ro-solution, exercise.reveal_solution=FALSE}
simulation_RO <- cohort_simulation(
  initial_probs = probs,
  parameters_UC = params_UC,
  parameters_TX = params_RO,
  horizon = 10, discount_rate = 0.035
)
ICER_RO <- calculate_ICER(simulation_RO)
ICER_RO
```

```{r cea-ro-check}
grade_this_code()
```

## Exercises {data-progressive=FALSE}

We can consider a couple of extensions to this simple model -- I'll leave these for exercises for you to try if you're interested.

### Alternative outcome measures

In some scenarios, such as when the new intervention is cheaper but results in fewer QALYs than the comparator, or (especially) when one intervention dominates the other, the ICER is a less useful statistic.

An alternative measure is the incremental net monetary benefit, defined as the 'value' of the QALYs gained (or lost) at a defined willingness-to-pay level, minus the incremental cost of the intervention. Try to write a function to calculate this outcome, similar to our current ICER function. (Hint: the function will need an additional argument to supply the willingness-to-pay level.)

### Individual-level simulation model

A very similar code structure can be used to implement the individual-level simulation model: Define a function (e.g. `individual_simulation()`) to run the model for both the comparator and intervention scenarios and return calculated Costs and QALYs; and another function to calculate the desired outcome measures from these results.

